Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Elemental concentrations such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021]. 
Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augsbergerDifferencesLeafPhenology2003].
Within tropical dry-deciduous species in Panama, the fraction of seasonally deciduous individuals increased toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. 
Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. 
For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. 
Leaf age affects vulnerability to heat stress and has a strong effect on metabolism [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Leaf hydraulic, stomatal, and boundary layer conductance are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), boundary layer conductance increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, because water supply often cannot meat the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].

@@

Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008]. 
Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008]. 
Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g. @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].

We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020]. 
Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018]. 
In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a]. 
@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous. 
Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. 
Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021]. 
Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Tree metabolism and growth and shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates, measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, @stephensonRateTreeCarbon2014]. 
NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

>>>>

Elemental concentrations such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021]. 
Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of nitrogen and phosphorus and foliar thermal tolerance.

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augspurgerDifferencesLeafPhenology2003].
Within tropical dry-deciduous species in Panama, the fraction of seasonally deciduous individuals increased toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. 
Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. 
For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. 
Leaf age affects vulnerability to heat stress and has a strong effect on metabolism [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Leaf hydraulic, stomatal, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].

@@

Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008]. 
Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008]. 
Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g. @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].

We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020]. 
Similarly, $T_{opt}$ of RuBP carboxylation  ($Vc_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018]. 
In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a]. 
@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous. 
Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. 
Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021]. 
Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates, measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006; @stephensonRateTreeCarbon2014]. 
NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
